Concurrent Programming in Harmony

### 2020-12-31 book/paper.tex
Fixing various things in the book

Mesa monitors provide another important and convenient option:
operation \texttt{notifyAll}
notifies \emph{all} threads that are waiting on a condition
instead of just one.
Having such an operation is not possible with Hoare monitors because using
Hoare monitors control
must be passed immediately to a thread that has been signaled, and that
can only be done if there is just one such thread.
\index{notifyAll}%
(aka \texttt{broadcast)})
\index{broadcast}%
lets all threads in that bedroom into the hall.

The so-called
``Mesa monitor semantics'' or ``Mesa condition variable'' semantics
have become more popular than Hoare monitor semantics, and have been
adopted by all major programming languages.
That said, few programming languages provide full support for monitors.
In Java, each object has a hidden lock \emph{and} a hidden condition variable
associated with it.
Methods declared with the \texttt{synchronized} keyword automatically
obtain the lock.  Java objects also support \texttt{wait}, \texttt{notify},
and \texttt{notifyAll}.
In addition, Java supports explicit allocations of locks
and condition variables.
In Python, locks and condition variables must be explicitly declared.
The \texttt{with} statement makes it easy to acquire and release a lock
for a section of code.
In C and C++, support for locks and condition variables is through libraries.

@@

Like C, Java, and Python, Harmony does not have built-in language
support for monitors, but provides library support through the
Harmony \texttt{synch} module.
\autoref{fig:cv} shows the implementation of
condition variables in the \texttt{synch} module.
\texttt{Condition}() creates a new condition variable.
It is represented by a dictionary containing
a bag of contexts of threads waiting on the condition variable.
(The \texttt{synchS} library instead uses a list of contexts.)

@@

To illustrate how Mesa condition variables are used in practice, we demonstrate using an implementation of reader/writer locks.
\autoref{fig:RWcv} shows the code.  \textit{mutex} is the shared lock
that protects the critical region.
(Again, there is only one ``gate.'')
There are two condition variables: readers wait on \textit{r\_cond} and
writers wait on \textit{w\_cond}.
The implementation also keeps track of the number of
readers and writers in the critical section.

Note that \texttt{wait} is always invoked within a \textbf{while}
loop that checks for the condition that the thread is waiting for.
It is \emph{imperative} that there is always a \textbf{while} loop
around any invocation of \texttt{wait} containing the negation of
the condition that the thread is waiting for.  Many implementation
of condition variables depend on this, and optimized implementations
of condition variables often allow so-called ``spurious wakeups,''
where \texttt{wait} may sometimes return even if the conditon
variable has not been notified.
In particular, one should always be able to replace \texttt{wait}
by \texttt{unlock} followed by \texttt{lock}.
This turns the solution into a busy-waiting one, inefficient but still
correct.

@@

\texttt{release\_wlock} executes \texttt{notifyAll}(?\textit{r\_cond})
as well as \texttt{notify}(?\textit{w\_cond)}.  Again, because we
do not keep track of the number of waiting readers or writers, we
have to conservatively assume that all waiting readers can enter,
or, alternatively, up to one waiting writer can enter.  So
\texttt{release\_wlock} wakes up all potential candidates.  There
are two things to note here.  First, unlike split binary semaphores
or Hoare monitors, where multiple waiting readers would have to be
signaled one at a time in a baton-passing fashion (see
\autoref{fig:RWsplitsema}), with Mesa monitors all readers are
awakened in one fell swoop using \texttt{notifyAll}.  Second, both
readers and writers are awakened---this is ok because both execute
\texttt{wait} within a \textbf{while} loop, re-checking the condition
that they are waiting for.  So if both type of threads are waiting,
either all the readers get to enter next or one of the writers gets
to enter next.

While Hoare moniters embody the split binary semaphore approach,
Mesa monitors are much closer to the busy-waiting approach.
The Mesa code then has to be careful to invoke \textbf{notify} or
\textbf{notifyAll} in the right places.
Much of the complexity of programming with Mesa condition variables
is in figuring out when to invoke \texttt{notify} and when to invoke
\texttt{notifyAll}.  As a rule of thumb: be conservative---it is
better to wake up too many threads than too few.  Waking up too
many threads may lead to some inefficiency, but waking up too few
may cause the application to get stuck.  Harmony can be particularly
helpful here, as it examines every corner case.  You can safely try
to replace each \texttt{notifyAll} with \texttt{notify} and see if
every possible execution of the the application still terminates.

@@

\section*{Exercises}
\begin{problems}
\item \label{ex:bbmesa} Implement a solution to the bounded buffer problem using Mesa condition
variables.
\item Implement a ``try lock'' module using Mesa condition variables
(see also \autoref{ex:trylock}).  It should
have the following API:
\begin{enumerate}
\item \textit{tl} = \texttt{TryLock}() \# \emph{create a try lock}
\item \texttt{acquire}(?\textit{tl}) \# \emph{acquire a try lock}
\item \texttt{tryAcquire}(?\textit{tl}) \# \emph{attempt to acquire a try lock}
\item \texttt{release}(?\textit{tl}) \# \emph{release a try lock}
\end{enumerate}
\noindent
\texttt{tryAcquire} should not wait.
Instead it should return \texttt{True} if the lock was successfully
acquired and \texttt{False} if the lock was not available.
% \item Implement semaphores using Mesa condition variables.
\item Write a new version of the GPU allocator in \autoref{ex:gpu}
using Mesa condition variables.
In this version,
a thread is allowed to allocate a set of GPUs and release a set of GPUs that it
has allocated.  Method \texttt{gpuAllocSet(n)} should block until $n$ GPUs are
available, but it should grant them as soon as they are available.
Method \texttt{gpuReleaseSet(s)} takes a set of GPU identifiers as argument.
A thread does not have to return all the GPUs it allocated at once.
\item The specification in the previous question makes the solution unfair.
Explain why this is so.  Then change the specification and the solution so that
it is fair.
\item \label{ex:qsort}
Bonus problem: \autoref{fig:qsort} shows an iterative implementation of the Qsort
algorithm, and \autoref{fig:qsorttest} an accompanying test program.
The array to be sorted is stored in shared variable \textit{textqs.arr}.
Another shared variable, \textit{testqs.todo}, contains the ranges of the
array that need to be sorted (initially the entire array).
Re-using as much of this code as you can, implement a parallel version of
this.  You should not have to change the methods \texttt{swap}, \texttt{partition},
or \texttt{sortrange} for this.  Create \texttt{NWORKERS} ``worker threads''
that should replace the \texttt{qsort} code.
Each worker loops until \textit{todo}
is empty and sorts the ranges that it finds until then.  The \texttt{main}
thread needs to wait until all workers are done.
\end{problems}

@@

Imagine five philosopers sitting around a table, each with a plate of food in
front of them and a fork between each two plates.  Each philosopher requires
two forks to eat.  To start eating, a philosopher first picks up the fork on
the left, then the fork on the right.  Each philosopher likes to take breaks
from eating to think for a while.  To do so, the philosopher puts down both
forks.  Each philosopher repeats this procedure.  Dijkstra had them repeating
this for ever, but for the purposes of this book, philosophers can leave
the table when they're not eating.

\autoref{fig:diners} implements the dining philosophers in Harmony, using a
thread for each philosopher and a lock for each fork.  If you
run it, Harmony complains that the execution may not terminate, with all five
threads being blocked trying to acquire the lock.

@@

\noindent
There are four conditions that must hold for deadlock to occur~\cite{CES71}:
\begin{enumerate}
\item \emph{Mutual Exclusion}: each resource can only be used by one thread at a time:
\item \emph{Hold and Wait}: each thread holds resources it already allocated while it
waits for other resources that it needs;
\item \emph{No Preemption}: Resources cannot be forcibly taken away from threads that
allocated them;
\item \emph{Circular Wait}: There exists a directed circular chain of threads, each waiting
to allocate a resource held by the next.
\end{enumerate}

Preventing deadlock thus means preventing that one of these conditions occurs.
However, mutual exclusion is not easily prevented (although, for some resources it is
possible, as demonstrated in \autoref{ch:nonblocking}).
Havender proposed the following techniques that avoid the remaining
three conditions~\cite{Havender68}:

@@

\section*{Exercises}
\begin{problems}
\item \label{ex:bank} \autoref{fig:bank} shows an implementation of a bank with various
accounts and transfers between those accounts.
Unfortunately, running the test reveals that it sometimes leaves unterminated
threads.  Can you fix the problem?
\item Add a method \texttt{total()} to the solution of the previous question
that computes the total over all balances.
It needs to obtain a lock on all accounts.  Make sure that
it cannot cause deadlock.
\item Implement an extra thread that checks, once, if the total of
the balances is the same as that at the beginning (using an assertion).
Is once enough to determine that no money gets lost or created, or should it do
so continually?
\end{problems}

@@

\begin{figure}
\begin{code}
\harmonysource{barrier}
\end{code}
\caption{\harmonylink{code/qbarrier.hny} Barrier synchronization with queues.}
\label{fig:barrier}
\end{figure}

>>>>

Mesa monitors also allow notifying multiple threads.
For example, a thread can invoke \texttt{notify()} twice---if there are
two or more threads waiting on the condition variable, two will be resumed.
Operation \texttt{notifyAll}
\index{notifyAll}%
(aka \texttt{broadcast)})
\index{broadcast}%
notifies \emph{all} threads that are waiting
on a condition.
Signaling multiple threads is not possible with Hoare monitors because with
Hoare monitors control
must be passed immediately to a thread that has been signaled, and that
can only be done if there is just one such thread.

The so-called
``Mesa monitor semantics'' or ``Mesa condition variable'' semantics
have become more popular than Hoare monitor semantics, and have been
adopted by all major programming languages.
That said, few programming languages provide full syntactical support
for monitors, instead opting to support monitor semantics through library
calls.
In Java, each object has a hidden lock \emph{and} a hidden condition variable
associated with it.
Methods declared with the \texttt{synchronized} keyword automatically
obtain the lock.  Java objects also support \texttt{wait}, \texttt{notify},
and \texttt{notifyAll}.
In addition, Java supports explicit allocations of locks
and condition variables.
In Python, locks and condition variables must be explicitly declared.
The \texttt{with} statement makes it easy to acquire and release a lock
for a section of code.
In C and C++, support for locks and condition variables is entirely
through libraries.

@@

Like C, Java, and Python, Harmony does not have syntactical
support for monitors. Harmony provides support for monitors
through the Harmony \texttt{synch} module.
\autoref{fig:cv} shows the implementation of
condition variables in the \texttt{synch} module.
\texttt{Condition}() creates a new condition variable.
It is represented by a dictionary containing
a bag of contexts of threads waiting on the condition variable.
(The \texttt{synchS} library instead uses a list of contexts.)

@@

To illustrate how Mesa condition variables are used in practice, we demonstrate using an implementation of reader/writer locks.
\autoref{fig:RWcv} shows the code.  \textit{mutex} is the shared lock
that protects the critical region.
There are two condition variables: readers wait on \textit{r\_cond} and
writers wait on \textit{w\_cond}.
The implementation also keeps track of the number of
readers and writers in the critical section.

Note that \texttt{wait} is always invoked within a \textbf{while}
loop that checks for the condition that the thread is waiting for.
It is \emph{imperative} that there is always a \textbf{while} loop
around any invocation of \texttt{wait} containing the negation of
the condition that the thread is waiting for.  Many implementation
of condition variables depend on this, and optimized implementations
of condition variables often allow so-called ``spurious wakeups,''
where \texttt{wait} may sometimes return even if the conditon
variable has not been notified.
As a rule of thumb, one should always be able to replace \texttt{wait}
by \texttt{unlock} followed by \texttt{lock}.
This turns the solution into a busy-waiting one, inefficient but still
correct.

@@

\texttt{release\_wlock} executes \texttt{notifyAll}(?\textit{r\_cond})
as well as \texttt{notify}(?\textit{w\_cond)}.  Because we
do not keep track of the number of waiting readers or writers, we
have to conservatively assume that all waiting readers can enter,
or, alternatively, up to one waiting writer can enter.  So
\texttt{release\_wlock} wakes up all potential candidates.  There
are two things to note here.  First, unlike split binary semaphores
or Hoare monitors, where multiple waiting readers would have to be
signaled one at a time in a baton-passing fashion (see
\autoref{fig:RWsplitsema}), with Mesa monitors all readers are
awakened in one fell swoop using \texttt{notifyAll}.  Second, both
readers and writers are awakened---this is ok because both execute
\texttt{wait} within a \textbf{while} loop, re-checking the condition
that they are waiting for.  So if both type of threads are waiting,
either all the readers get to enter next or one of the writers gets
to enter next.  (If you want to prevent waking up both readers and
a writer, then you can keep track of how many threads are waiting
in the code.)

While Hoare moniters embody the split binary semaphore approach,
Mesa monitors are much closer to the busy-waiting approach.
The Mesa code then has to be careful to invoke \textbf{notify} or
\textbf{notifyAll} in the right places.
Much of the complexity of programming with Mesa condition variables
is in figuring out when to invoke \texttt{notify} and when to invoke
\texttt{notifyAll}.  As a rule of thumb: be conservative---it is
better to wake up too many threads than too few.  Waking up too
many threads may lead to some inefficiency, but waking up too few
may cause the application to get stuck.  Harmony can be particularly
helpful here, as it examines every corner case.  You can safely try
to replace each \texttt{notifyAll} with \texttt{notify} and see if
every possible execution of the application still terminates.

@@

\section*{Exercises}
\begin{problems}
\item \label{ex:bbmesa} Implement a solution to the bounded buffer problem using Mesa condition
variables.
\item Implement a ``try lock'' module using Mesa condition variables
(see also \autoref{ex:trylock}).  It should
have the following API:
\begin{enumerate}
\item \textit{tl} = \texttt{TryLock}() \# \emph{create a try lock}
\item \texttt{acquire}(?\textit{tl}) \# \emph{acquire a try lock}
\item \texttt{tryAcquire}(?\textit{tl}) \# \emph{attempt to acquire a try lock}
\item \texttt{release}(?\textit{tl}) \# \emph{release a try lock}
\end{enumerate}
\noindent
\texttt{tryAcquire} should not wait.
Instead it should return \texttt{True} if the lock was successfully
acquired and \texttt{False} if the lock was not available.
% \item Implement semaphores using Mesa condition variables.
\item Write a new version of the GPU allocator in \autoref{ex:gpu}
using Mesa condition variables.
In this version,
a thread is allowed to allocate a set of GPUs and release a set of GPUs that it
has allocated.  Method \texttt{gpuAllocSet(n)} should block until $n$ GPUs are
available, but it should grant them as soon as they are available.
Method \texttt{gpuReleaseSet(s)} takes a set of GPU identifiers as argument.
A thread does not have to return all the GPUs it allocated at once.
(You may want to try implementing this with Split Binary Semaphores.  It is not as easy.)
\item The specification in the previous question makes the solution unfair.
Explain why this is so.  Then change the specification and the solution so that
it is fair.
\item \label{ex:qsort}
Bonus problem: \autoref{fig:qsort} shows an iterative implementation of the Qsort
algorithm, and \autoref{fig:qsorttest} an accompanying test program.
The array to be sorted is stored in shared variable \textit{textqs.arr}.
Another shared variable, \textit{testqs.todo}, contains the ranges of the
array that need to be sorted (initially the entire array).
Re-using as much of this code as you can, implement a parallel version of
this.  You should not have to change the methods \texttt{swap}, \texttt{partition},
or \texttt{sortrange} for this.  Create \texttt{NWORKERS} ``worker threads''
that should replace the \texttt{qsort} code.
Each worker loops until \textit{todo}
is empty and sorts the ranges that it finds until then.  The \texttt{main}
thread needs to wait until all workers are done.
\end{problems}

@@

Imagine five philosopers sitting around a table, each with a plate of food in
front of them and a fork between each two plates.  Each philosopher requires
two forks to eat.  To start eating, a philosopher first picks up the fork on
the left, then the fork on the right.  Each philosopher likes to take breaks
from eating to think for a while.  To do so, the philosopher puts down both
forks.  Each philosopher repeats this procedure.  Dijkstra had them repeating
this for ever, but for the purposes of this book, philosophers can leave
the table when they want to leave.

\autoref{fig:diners} implements the dining philosophers in Harmony, using a
thread for each philosopher and a lock for each fork.  If you
run it, Harmony complains that the execution may not be able to terminate,
with all five threads being blocked trying to acquire the lock.

@@

\noindent
There are four conditions that must hold for deadlock to occur~\cite{CES71}:
\begin{enumerate}
\item \emph{Mutual Exclusion}: each resource can only be used by one thread at a time:
\item \emph{Hold and Wait}: each thread holds resources it already allocated while it
waits for other resources that it needs;
\item \emph{No Preemption}: resources cannot be forcibly taken away from threads that
allocated them;
\item \emph{Circular Wait}: there exists a directed circular chain of threads, each waiting
to allocate a resource held by the next.
\end{enumerate}

Preventing deadlock thus means preventing that one of these conditions occurs.
However, mutual exclusion is not easily prevented in general
(although, for some resources it is
possible, as demonstrated in \autoref{ch:nonblocking}).
Havender proposed the following techniques that avoid the remaining
three conditions~\cite{Havender68}:

@@

\section*{Exercises}
\begin{problems}
\item \label{ex:bank} \autoref{fig:bank} shows an implementation of a bank with various
accounts and transfers between those accounts.
Unfortunately, running the test reveals that it sometimes leaves unterminated
threads.  Can you fix the problem?
\item Add a method \texttt{total()} to the solution of the previous question
that computes the total over all balances.
It needs to obtain a lock on all accounts.  Make sure that
it cannot cause deadlock.
\item Add an invariant that checks that the total of the balances never
changes.  Note that the invariant only holds if none of the locks are
held.
\end{problems}

@@

\begin{figure}
\begin{code}
\harmonysource{qbarrier}
\end{code}
\caption{\harmonylink{code/qbarrier.hny} Barrier synchronization with queues.}
\label{fig:barrier}
\end{figure}
